---
title: 'Smolt Abundance Estimates: MY 2017, Minam trap'
output:
  pdf_document:
    toc: yes
  html_document:
    toc: yes
    toc_float: yes
    toc_collapsed: yes
---

*This reported was generated on `r Sys.Date()`.*
```{r libraries, echo=FALSE, warning=FALSE, message=FALSE}
# This section needs to be run only once per session, to load the libraries
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE)
library(knitr)
library(kableExtra)
```

```{r Setup, echo=FALSE, warning=FALSE, message=FALSE}
my.txt <- "my17"
site.cur <- "MIN"

source(paste("code//site_setup//setup", my.txt, site.cur, "abundance.R", sep="_"))  #reads in the site-specific setup file, which also brings in general setup and custom functions.

```


This report documents decisions used to develop estimates of smolt abundance at the Lostine trap during migratory year (MY) 2017, including both fall (early migrants) and spring (late migrants) seasons. At each step, results are compared with the methods and results from the previous abundance analysis for this site, as performed by Brad Garner and as reported in the ELH annual report.  This previous analysis is sometimes referred to as "BG", for Brad Garner. ("BP" = "Brad Power", who took over the abundance analysis after BG left, for the MY2018 and MY2019 reports.)

Site Code: **`r site.cur`**

Analysis options:

- Was there any subsampling for this data set? **`r sub.sample`**

- Should *catch* (aka, total new) numbers from INT days be included? **`r include.INT.catch`**

- Which abundance estimator is used for simple abundance? **`r abund.est`**

# PROCESS TRAP LOG
```{r Initial_Processing}
# Run first part of data processing, before the first custom mod
source("code//1a_AbundancePrep_INT.R")

# Sort out fish cols for easier processing
# Create a version of log.te that doesn't have all the fish columns in the way 
  fish.cols <- colnames(log)[17:57]
  
# Assign an automatic j.wk to confirm trap log versions
  log.te$j.wk.auto <- ceiling(log.te$jday/7)
```

*Do the j.wks used in the trap log match the automatically assigned j.wks?*
```{r, echo=TRUE}
sum(log.te$j.wk != log.te$j.wk.auto)  #should be zero.
```



*Which days of trapping have usable fish data?*

Trap modes:
```{r, echo=FALSE, eval=TRUE}
table(log$trapmode)
```
The answer to this question differs for catch vs mark vs recapture numbers. 

### Trap Mode = INT (Interrupted)
- catch numbers are included
- recap numbers, and mark (release) numbers from the previous day, are excluded from TE estimates

Standard ELH analysis protocol is that **mark** and **recapture** numbers associated with INT days (ie, the recaptures on that day and the releases from the day prior) numbers should be excluded from trap efficiency calculations. This is explicitly specified in the protocols for trap log preparation for abundance, and can be seen in most of the analysis working-spreadsheets from recent years (both BP's and BG's). I confirmed that BG did remove the appropriate marks/recaps from INT days for the Minam MY 2017 analysis (for both CHS and STS). 

Unlike marks and recaps, trap **catch** numbers from INT days are always included in the total population estimate. Although this feels inconsistent, it is true that the simple population estimates that ELH has been reporting were already minimum values only, so including catch from more days, even if they are only partial days, will presumably only get the estimate closer to the true abundance. I confirmed that protocol in past analyses was indeed to include catch from INT days by reviewing some of the analysis working-spreadsheets, where catch from INT days was always included, even when the associated mark and recap numbers were successfully excluded. 

Additionally, the trap log preparation protocol written/revised by BG suggests that the analyst should examine the INT days, and, depending on total revolutions and fish catch numbers (relative to nearby CON days), decide whether any of the INT days can be considered as having usable data for purposes of mark/recapture. 

Examining the INT days from Minam MY17:
```{r, echo=FALSE}
filter(select(log, season, date, jday, j.wk, gage, trapmode, revolutions, chs.new, sts.new), trapmode=="INT") %>% #, comments
  kable() %>%
  kable_styling(bootstrap_options = ("striped"), full_width=TRUE, position="left", font_size=10)

```
Notes on the INT days:

- 10/16: no; CHS catch relatively low.

- 3/16: no; no fish data collected.

- 3/17: no; revs very low, chs numbers on the low side.

- 3/19: no; no fish data collected.

- 3/31: YES; no cmt in trap log about what caused INT, but revs are higher than either neighboring day, and catch numbers look reasonably consistent. Additionally, BG also chose to include this INT day (it is not highlighted in red, and he did remove the corresponding mark/recap counts).

- 4/4: no; revs and catch very low.

- 5/5: possible; revs pretty close to previous day, chs catch pretty close to previous day, sts catch relatively high. However, BG excluded this day, so I will do the same.

- 5/6: no; revs and catch are low.

- 5/10: no; revs significantly lower (9.5k, vs 14k on neighboring days.)

- 5/12: YES; revs higher than either previous day, fish catch numbers consistent (though not high), and BG also chose to include this INT day.

- 5/24: no, revs half the previous day.

Conclusion: Of these INT dates, 3/31 and 5/12 look good enough to include in the mark/recapture TE analysis.  This matches BG's prior analysis.  

```{r, echo=TRUE}
INT.mods

```


### Trap Mode = SUB (Subsample)
No subsampling at Minam! 

### Trap Mode = NR (not running)
As above, standard protocol is that TE releases prior to stopping the trap (usually unexpectedly; eg, fish were released, then trap stopped due to excessive debris) should be excluded from the TE calculations. There were no recorded instances of this situation at Minam for MY 2017. 
```{r, echo=FALSE, eval=FALSE}
filter(select(log, date, jday, j.wk, trapmode, startstop, chs.te.all, sts.te.all), startstop=="STP", chs.te.all>0 | sts.te.all>0) %>%
  kable() %>%
  kable_styling(bootstrap_options = ("striped"), full_width=FALSE, position="left", font_size=10)

```

Data check: are there any instances of [mark.data.good]=TRUE when [startstop]=STP? (generally, this should not happen) -> NO, not in this data set. (The one <NA> is from the last trapping day of the season.)7
```{r}
table(log.te$startstop, log.te$mark.data.good)
#filter(select(log.te, -fish.cols), mark.data.good==TRUE, startstop=="STP") #this one is okay because next day is CAL.
```


### Total counts
```{r source_1b_AbundancePrep, message=FALSE, warning=FALSE}
# Given custom mods to INT days (etc), run calculation of CMR counts.
source("code//1b_AbundancePrep_CMR.R")

```

With the above parameters, total catch/mark/recapture counts for the season:
```{r season_CMR_counts, echo=FALSE, message=FALSE, warning=FALSE}
season.cts <- cmr.counts %>%
  group_by(myear, site, species, season) %>%
  summarize(catch = sum(catch),
            mark = sum(mark),
            rec = sum(rec))
season.cts <- mutate(season.cts, te.all = round(rec/mark,3))

season.counts.tb <- kable(season.cts) %>%
  kable_styling(bootstrap_options = ("striped"), full_width=FALSE, position="left", font_size=10)

season.counts.tb
```

Compare the previous, BG version
```{r, echo=FALSE}
# Manually recreate BG's table for LOS MY17
season.cts.bg <- season.cts
season.cts.bg$catch <- c(13795, 598, 1040, 474)
season.cts.bg$mark  <- c(914, 436, 495, 323)
season.cts.bg$rec   <- c(235, 20, 37, 7)
season.cts.bg <- mutate(season.cts.bg, te.all = round(rec/mark,3))

season.cts.bg.tb <- kable(season.cts.bg) %>%
  kable_styling(bootstrap_options = ("striped"), full_width=FALSE, position="left", font_size=10)

season.cts.bg.tb

```
Discrepancies are due to:

- CHS fall marks: I have 245, BG has 235 -> due to 10 OTE exclusions by BG, I will deal with OTE below.
- STS fall marks: I have 496, BG has 495; looks like this is due to an addition error in the trap log [Total TE] column ([AY$132]), so my number is correct.
- STS fall recaps: I have 38, BG has 37 -> due to another [Total TE RE] addition error ([BD$123]), so my number is correct. 

## Stratification
Stratification approach used in previous years seems to be somewhat subjective. It largely follows the >=10 rule, but BG's modifications to the trap log preparation protocol suggest that he aimed to combine "where possible". From the protocol: 
*We need at least 10 recaptures with every trap efficiency estimate.  Group weeks whenever weekly recaptures are less than 10.  Also, group weeks where water temperature and discharge are similar as these are primary influences on fish movement and trap capture efficiency. (Note: it is preferable to group weeks rather than not, this allows more OTE data to be used).*
(Of course, if this logic were taken to its fullest extent, then we would not stratify at all.)

The general goal in this re-analysis of previous MY data is to use BG's stratification, except where there is some specific reason to modify it.

**Comments on specific stratification decisions for each species/season**

- CHS fall: good recaps this season.  Stratifications are identical; mostly individual weeks, except with the first two and then the last two TE weeks pooled.  

- CHS spring: only 20 total recaps for the season. Identical stratifications, first three weeks in one stratum and remaining weeks all a second stratum. 

- STS fall: 37 total recaps for the seasons. The two stratifications are close, with the exception of weeks 40 (2 recaps) and 41 (1 recap).  A standard stratification puts these weeks in stratum b (resulting in recap numbers of 11, 14, and 13 for the three strata), while BG put them in stratum a (resulting in recap numbers of 14, 11, and 13).  Looking at the gage data supports BG's custom stratificaiton: discharge suddenly jumps way up at the beginning of week 42 (conveniently right at the same time as the start of a new j.week) -> see plot below.

- STS spring: only 7 total recaps for the entire season, so no stratification.  There were 323 fish marked and released, so STS trap efficiency for this season was very low (~ 2%).

```{r, echo=FALSE, warning=FALSE, message=FALSE}
print(ggplot(data=log.te[log.te$date<"2016-12-01",], aes(x=date)) + 
  geom_point(aes(y=as.numeric(gage.cfs))) +
  geom_vline(aes(xintercept=min(log.te$date[log.te$j.wk==42])), lty=2) +
  ylab("discharge (cfs)") +
  labs(title="Minam river discharge, fall MY17",
             subtitle = ("Dashed line shows the start of TE week 42, and the start of BG's stratum b")) )

filter(select(log.te, date, j.wk, gage.cfs, temp), date>"2016-10-09", date<"2016-10-18") %>%
  kable() %>%
  kable_styling(bootstrap_options = ("striped"), full_width=FALSE, position="left", font_size=10)
```

Conclusion: use BG's previous stratification, which differs from a standard stratification only for STS weeks 40 and 41, as described above. No other custom modifications needed. 
```{r show_strat.mods, echo=TRUE}
use.prv.strata
strat.mods
```

```{r source_1c_AbundancePrep, warning=FALSE, message=FALSE}
source("code//1c_AbundancePrep_StratificationOTE.R")
```

## Overdue TE fish
There were `r sum(ote.str$count)` fish noted as OTE in the MIN MY17 trap log, primarily CHS and all but one PIT marked (meaning that release week is known exactly).  Of these, `r sum(ote.str$count[ote.str$ote==TRUE])` fish were confirmed OTE, i.e., they were released and recaptured in different strata, and therefore they were removed from both release (mark) and recapture counts for the TE analysis.  

```{r}
ote.str %>%
  arrange(date) %>%
  kable() %>%
  kable_styling(bootstrap_options = ("striped"), full_width=FALSE, position="left", font_size=10)
```
Note that BG removed confirmed OTE fish from recap counts, **but not from the corresponding mark counts**. Therefore my season total marks for fall CHS (904) is 10 less than BG's season total (914).

Revised season totals, now incorporating OTE removals:
```{r revised_season_cts}
season.cts.rev %>%
  kable() %>%
  kable_styling(bootstrap_options = ("striped"), full_width=FALSE, position="left", font_size=10)
```


## Subsampling
There is no subsampling at Minam. No custom mods needed for subsample date assignments. 
```{r source_2_Calendar}
source("code//2_FullCalendar.R")
```

```{r source_2_SubsamplePrep}
# Run subsampling prep code only if subsampling occurred at this site/year
if (sub.sample == TRUE) {
  source("code//3_SubsamplePrep.R")
}
```


# ABUNDANCE ESTIMATION
For comparison, abundance here is calculated in two different ways. 

(1) Simple estimate using TE estimates by stratum (with bootstrapped variance est.) There is also a version of the simple abundance estimate derived from the bootstrap routine (ELH project has traditionally been using this value for the reported abundance estimates).  

(2) Estimated based on a GAM that interpolates for missed trapping days; as implemented by Duarte and Petersen (rev. 2020).
 
Method 1 is a minimum estimate, which ignores dates when the trap was not running.
Method 2 uses the surrounding data to interpolate estimates for the missed trapping dates. 

The immediately following diagnostic plots (if any) apply to the GAM-fitting in Method 2. There should be four sets of plots, one set each for 

- CHS fall
- CHS spring
- STS fall
- STS spring 

```{r source_abundance, warning=FALSE, message=FALSE}
# First clean up environment, except for run-specific variables
  rm(list=ls()[!ls() %in% c("my.txt", "my.cur", "site.cur")])
# ...and rerun site setup
  source(paste("code//site_setup//setup", my.txt, site.cur, "abundance.R", sep="_"))

# Set to FALSE to suppress printing the diagnostic plots
  diagnostics <- FALSE

# Now source abundance calculation script (which will also source boot and gam routines)
  source(paste(dir_code, "4_CalculateAbundance.R", sep=""))
```

### 1. Simple/minimum estimates
Apply the TE estimates for each stratum (calculated using the `r abund.est` estimator) to estimate the total number of fish moving past the trap that each stratum (j). Because no attempt is made to account for days when the trap was not operating, this approach calculates a **minimum** estimate for the total number migrants. 

Essentially, this method is how the past abundance estimates reported by ELH have been calculated, although it seems that at least recently staff have actually been reporting the abundance estimates calculated by the bootstrap routine (which are not identical to the simple approach described above, but in practice the two methods produce very similar values).  For MY2017, it appears that BG applied the GAM routine to calculate abundance, but only for Catherine Creek; for the other sites, he used the standard simple/bootstrap estimates of abundance. 

Using the Chapman estimator [(R-1)/(M-1)] rather than the simple Petersen estimator [R/M] leads to larger TE values, which leads to lower estimated abundance. Chapman weekly abundance estimates for Lostine were reduced by about 5% on average (range 1.75 - 9.09) relative to Petersen estimates. The magnitude of this effect depends on sample size, primarily the number of recaps: fewer recaps = greater difference between Chapman vs Petersen estimates = greater reduction in abundance (relative to Petersen).   

Variance for the simple abundance estimates is calculated using a one-sample bootstrap method as described in Thedinga et al.  Note that the current (June 2020) version of the bootstrap routine mixes (I think this is an oversight) elements of both Petersen and Chapman estimators: The initial TE and abundance estimates are calculated using Chapman, but then when generating the bootstrapped vector of estimated TE values, it uses simple Petersen estimates.  The net effect is that the abundance estimates reported by the boot routine are much closer to the simple Petersen estimate than to the simple Chapman estimate.  This inconsistency is something I will fix after talking with Duarte and Peterson. Not sure how it impacts variance estimates.

```{r SimpleAbundance, echo=FALSE}
merge(select(simp.byseas, myear, site, species, season, est.simp),
      select(boot.byseas, myear, site, species, season, est.boot, se.boot),
      by=c("myear", "site", "species", "season"), all=TRUE) %>%
  kable(digits=0) %>%
  kable_styling(bootstrap_options = ("striped"), full_width=FALSE, position="left", font_size=10)
```

Compare the simple values calculated by BG:
```{r BG_SimpAbund}
# Manually compile BG's numbers in the order chs.fall, chs.spring, sts.fall, sts.spring
bg.abund <- data.frame(species = c("chs", "chs", "sts", "sts"), 
                       season = c("fall", "spring", "fall", "spring"),
                       abund = c(62430, 13619, 14404, 22418), 
                       ci.95 = c(17558, 4672, 5944, 8129) ) %>%
  mutate(se = ci.95/1.96)

bg.abund %>%
  kable(digits=0) %>%
  kable_styling(bootstrap_options = ("striped"), full_width=FALSE, position="left", font_size=10)

```
My recalculated abundance estimates are very close to BG's numbers.  The spring season calculations used identical season totals, and identical stratification, so theoretically our results ought to be identical. Presumably the minor discrepancies in our results are due to different bootstrap runs (a randomization process). For the fall season estimates, due to slight differences in mark and recap numbers as described above, my TE estimates are slightly higher, and therefore my total abundance estimates slightly lower, than BG's versions.      

### 2. Interpolated estimates using a GAM
Migrant abundance was also estimated using a general additive model to smooth patterns in migrant numbers and to interpolate for missed trapping days. The following results were calculated using a routine created by Duarte and Petersen, and revised in Feb 2020 to improve the calculation of aggregate standard errors. See above for diagnostic plots associated with model fits. 

In the plots below, the daily estimated abundance values (daily catch expanded by TE value for the stratum) are plotted as grey (or green) points, while the GAM daily estimates are shown as red lines; the pale red ribbons show the the confidence intervals associated with each daily estimate. Thick black bars at the base of each plot indicate dates when the trap was not operating. 

```{r Plot_GAM, echo=FALSE, warning=FALSE, message=FALSE, fig.width=10}
# Create an index to visually show when the trap was not operating
  calendar.gam$gap.ind <- NA
    calendar.gam$gap.ind[calendar.gam$fish.data==FALSE] <- -2

# Create an index to mark SUB vs non-SUB days
  calendar.gam$mode.ind <- "CON"
    calendar.gam$mode.ind[calendar.gam$trapmode=="SUB"] <- "SUB"
    
# Plot daily estimated values vs measured catch
basic <- ggplot(data=calendar.gam, aes(x=date)) +
  geom_line(aes(y=catch/(rec/mark)), color="dark grey") +  #daily simple estimate
  geom_point(aes(y=catch/(rec/mark), color=mode.ind, shape=mode.ind)) +  #daily simple estimate
  geom_line(aes(y=est.gam), color="red") +  #interpolated
  geom_line(aes(y=gap.ind), color="black", size=2.5) +  #Clearly show when trap was not operatin
  geom_ribbon(aes(ymin=ci.low, ymax=ci.high), alpha=0.1, fill="red") + #Band representing std error
  facet_grid(species ~ season, scales="free") +
  xlab("Date") + ylab("Abundance") +
  scale_color_manual(values = c("#666666", "#009900"), 
                     name="Trap\nmode") +
  scale_shape_discrete(name="Trap\nmode") +
  theme(panel.grid.minor=element_blank()) + 
  theme_bw() 

basic
```

Zooming in on the lower portion of each plot, and adding additional unnec. text to avoid line break:
```{r Plot_GAM_zoom, echo=FALSE, warning=FALSE, message=FALSE, fig.width=10}
# Same as above but zooming in on lower portion
basic +
  coord_cartesian(ylim=c(0, 2000))

```

These abundance results look good. Modeled values follow measured values appropriately.  It helps that there are very few missed trapping days to worry about in this data set. Confidence intervals appear to get wider during non-operation periods, as appropriate.   

However, similar to results from the Lostine MY17 analysis, here the aggregated standard error values associated with these estimates seems very low, less than 1/10 the bootstrap SE estimates. In fact, they are closer to 1/100 the bootstrapped versions. This seems implausible.  
```{r GAM_results}
# Display a basic table of results
gam.byseas %>%
  kable(digits=0) %>%
  kable_styling(bootstrap_options = ("striped"), full_width=FALSE, position="left", font_size=10)
```

*Remaining questions about GAM results:* 

- Why are aggregate estimated standard errors for total abundance so low? What might be a better way to calculate this?


### 3. Compare abundance results
```{r Compare_Abund_results}
all.byseas[, 1:14] %>%
  kable(digits=0) %>%
  kable_styling(bootstrap_options = ("striped"), full_width=FALSE, position="left", font_size=10)
```

**Missed fish:**
How many fish were missed during non-operational periods? (According to GAM estimates.) 
```{r, warning=FALSE, message=FALSE}
missed <- calendar.gam %>%
  filter(fish.data==FALSE) %>%
  group_by(species, season) %>%
  summarize(fish.missed = sum(est.gam),
            days.missed = n())

missed <- merge(gam.byseas, missed, by=c("species", "season"), all=TRUE)
missed <- merge(missed, simp.byseas[, 1:5], 
                by=c("myear", "site", "species", "season"), all=TRUE)
missed <- mutate(missed, prop.missed = round(fish.missed/est.gam, 2))

  all.days <- calendar.gam %>%
    group_by(species, season) %>%
    summarize(n.days = n())

missed <- merge(missed, all.days, by=c("species", "season"), all=TRUE)

missed %>%
  select(myear, site, species, season, n.days, days.missed, 
         est.simp, est.gam, fish.missed, prop.missed) %>% 
  kable() %>%
  kable_styling(bootstrap_options = ("striped"), full_width=FALSE, position="left", font_size=10)

rm(all.days)
```
Relatively few days were missed this season, but they fell during periods of high catch, resulting in large numbers of missed fish.  

In the fall, the two missed days in November should be trivial; but there was a third missed day (which does not show up in the plot as a black bar, because it was only a single missed day) on October 16, right in the middle of the greatest peaks in movement for both CHS and STS.

```{r}
filter(calendar.gam, date>"2016-10-12", date<"2016-10-20") %>%
  select(myear, site, species, season, date, trapmode, catch, ,mark, rec, est.gam) %>%
  kable %>%
  kable_styling(bootstrap_options = ("striped"), full_width=FALSE, position="left", font_size=10)

```
Based on a simple interpolation, we would estimate simple abundances of `r round(mean(c(2304, 1557))/(18/94), 0)` (CHS) and `r round(mean(c(167, 169))/(11/160), 0)` (STS), which are reasonably close to the GAM-estimated versions. So the estimate of 18% of abundance missed while the trap was not operating seems plausible.  

In the spring, the missed days in mid-late March (3/16-3/17 and 3/19-3/20) also occurred during periods of peak movement for both CHS and STS. Additionally, the run of three missed days in early May (5/5-5/7) followed several days of relatively high STS catch and estimated abundance. About 1/3 of both CHS and STS abundance was missed during the spring.     

**Percent migrating early:** 
What proportion of the total number of migrants for this MY migrated past the trap in fall (early) vs in spring (late)?
```{r pct_mig_early}
totals <- all.byseas %>%
  group_by(species) %>%
  summarize(tot.simp= sum(est.simp),
            tot.gam = sum(est.gam))

totals <- merge(all.byseas, totals, by="species", all=TRUE)
totals <- mutate(totals, 
                 prop.simp = round(est.simp/tot.simp, 2),
                 prop.gam = round(est.gam/tot.gam, 2) )

select(totals, myear, site, species, season, 
       est.simp, prop.simp, 
       est.gam, prop.gam ) %>%
  kable() %>%
  kable_styling(bootstrap_options = ("striped"), full_width=FALSE, position="left", font_size=10)
```
For Chinook, about 80% of the total migration left in the fall; for steelhead, by contrast, the majority of migrants (~60%) left in the spring. Proportions of early vs. late migrants are very similar between simple vs. GAM abundance estimates. 


# MEDIAN MIGRATION DATE
For comparison, I estimated median migration date using three different methods, described below. I also manually brought in BG's values for comparison. 

- catch: based only on the numbers of fish caught in the trap. Note that for this analysis, I did use expanded catch estimates for subsample days. 

- exp.catch: based on daily catch numbers expanded by TE for the stratum. This method is equivalent to the simple abundance estimates. 

- gam: based on daily abundance values as estimated by the GAM routine. 

- bg: median dates as reported by BG. BG did not leave any documentation of his methods, but as far as I have been able to recreate, he calculated estimated abundance by TE week (so, using expanded catch, then summed by week), and then calculated median date as if all fish passing during that week had passed on the first day of the TE week. So in general, BG's dates are likely to be similar to my method (2), but a bit earlier, due to the inherent strong bias toward the beginning of the TE week. 

```{r calculate_median_dates}
# Bring in BG's values
  bg <- c("2016-10-14", "2017-03-19", "2016-10-14", "2017-04-09")
  median.dates <- cbind(median.dates, bg)
  
  rm(bg)

# Display the comparison table
  median.dates %>%
    kable() %>%
    kable_styling(bootstrap_options = ("striped"), full_width=FALSE, position="left", font_size=10)

```

To match the current standard reporting of a simple abundance estimate, I will save the analogous, expanded catch version of the median migration date for official reporting. 

*FINIS*